---
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
---
# Import/Export
Objective:  
- Learn how to import (read.csv(), read.delim()) and export tabular data (write.csv(), write.delim())
- Learn how to plot graphs using base R  
   
We will also cover: 
- use data sets provided by R
- how to create a new directory/folder using dir.create()  
- open several graphics windows  
- save plots to file  
   
## Export Tables
** Tips for Preparing Data**

- first row = column headers (variables) and first column = row names (observations) 
Row and column names  
- should be unique  
- shouldn't begin with a number  
- no special symbols (except underscore)  
- avoid blank spaces  
- avoid blank rows and comments  
- use the four digit format for date: Good: 01/01/2016. Bad: 01/01/16  
  
In Excel, save your file into .txt (tab-delimited text file) or .csv (comma separated value file) format.  
  
**Working with txt and csv files**
- R has built-in datasets that we may use (see datasets with data() command)
```{r}
# Call data into environment with data("dataset")
# Loading DNase data
data("DNase")
# Get more info using ?
# ?DNase
# See first 10 rows with head()
head(DNase, n = 10) 
# basic stats with summary()
summary(DNase)
# unique values of a vector with unique()
unique(DNase$Run)
```

**R base functions for exporting/writing/saving data:**
* write.table() - exporting any tabular data to ".txt" file
* write.csv() - comma separated values,".csv" file

variations when comma (,) is used as decimal points instead of periods (.): write.csv2()  
 
Arguments:  
- file = absolute/local file paths with name of file
- sep = field seperators (sep argument) - character(s), controls the way splits a data table into fields/cells  
- header = logical for column names  
- row.names = vector of row names or number/name of column which contains the row names  
```{r}
# Create a Data folder
dir.create("Data")
# Write data to txt file: tab separated values
# sep = "\t"
write.table(DNase, file = "Data/DNase.txt", sep = "\t", row.names = FALSE) #"\t" means "tab"

# Write data to csv files:
# decimal point = "." and value separators = comma (",")
write.csv(DNase, file = "Data/DNase.csv")

# Write data to csv files:
# decimal point = comma (",") and value separators = semicolon (";")
# write.csv2(DNase, file = "DNase.csv")
```

## Import Tables
**R base functions for importing data**

* read.table() - reading any tabular data  
* read.delim() - reading tab-separated value files (.txt)  
* read.csv() - comma separated value files (.csv)  
  
variations when comma (,) is used as decimal points: read.delim2(), read.csv2()

- file = absolute/local file paths with name of file, file.choose() to choose interactively, internet address e.g. "http://www.sthda.com/upload/boxplot_format.txt"  
```{r}
# Read tabular data into R
# file.choose() allows you to interactively pick a data file
# df <- read.table(file.choose(), header = FALSE, sep = "\t", dec = ".")
# Read "comma separated value" files (".csv")
filename <- "Data/DNase.csv"
df <- read.csv(filename, header = TRUE)
# Read TAB delimited files
df2 <- read.delim(file = "Data/DNase.txt", header = TRUE, sep = "\t", dec = ".")
# Note to save into an object, we must assign to variable!
```

**Working with Excel files**

* xlsx package  
* readxl package  
 
Note: install libraries/packages in R using install.packages("package_name"). Load intro environment before using with library(package_name)  

Writing Data From R to Excel Files (xls|xlsx)
```{r}
# # install.packages("openxlsx")
# 
# library("openxlsx") # or require("openxlsx")
# # Write the first 10 rows in a new workbook
# write.xlsx(DNase[1:10,], file = "Data/DNase.xlsx", sheetName = "DNase 1", append = FALSE)
# # Add a second data set in a new worksheet with first and third column
# write.xlsx(DNase[,c(1,3)], file = "Data/DNase.xlsx", sheetName = "DNase 2", append = TRUE)
```
**Reading data From Excel Files (xls|xlsx) into R**
```{r}
# # Use openxlsx package
# my_data <- read.xlsx("Data/DNase.xlsx", sheetIndex = 1) 
```

**Using R data format: RDATA and RDS**

- Save/load variables, user-made functions and data
Saving objects
- Save one object to a file: saveRDS(object, file)
- Save multiple objects to a file: save(data1, data2, file)
- Save your entire workspace (all objects): save.image()
Loading objects
- readRDS(rds_file), load(RData_file)
```{r}
# # Save a single object to a file
# saveRDS(DNase, "Data/DNase.rds")
# # Restore it under a different name
# my_data <- readRDS("Data/DNase.rds")
# # Save multiple objects
# save(DNase, my_data, file = "Data/data.RData")
# # To load the data again
# load("Data/data.RData")
# # Saving and restoring your entire workspace:
# # Save your workspace
# save.image(file = "Data/my_work_space.RData")
# # Load the workspace again
# load("Data/my_work_space.RData")
```

**Consider using readr package**

* Reading lines from a file: read_lines()  
* readr functions for writing data: write_tsv(), write_csv()  
* readr for reading/writing txt|csv files: read_tsv(), read_csv(), etc  
  
## Graphing with base R
- Now that we can import our data, we can learn how to plot it. 
- The purpose of base R plots is to analyze our data, not as elegant for publication purposes  
  
**plot()**

- generic function, meaning it can work with different types of objects  
- plot() function can be used to plot 2 variables (of equal length)
    plot(x, y, type=...)
    Arguments
        x and y = the coordinates of points to plot
        type = the type of graph to create, possible values:
          "p" - points - scatter plot  
          "l" - lines  
          "b" - both points and lines  
          "c" - empty points joined by lines  
          "o" - overplotted points and lines  
          "s" and "S" - stair steps  
          "h" - histogram-like vertical lines  
          "n" - does not produce any points or lines  
        main = Title for plot  
        xlab = Title for x axis   
        ylab = Title for y axis  
        
- You can specify fonts, colors, line styles, axes, reference lines, etc. by adding graphical parameters (cex, col, lwd, etc)
Read more: https://www.statmethods.net/advgraphs/parameters.html

- Combine plots using par()  
Read more: https://www.statmethods.net/advgraphs/layout.html 
```{r}
# You can plot all pairs of variables in one graph
plot(DNase)
# You can also specify vectors to plot
conc <- DNase$conc; dens <- DNase$density
plot(x = conc, y = dens, type ="p") 
plot(x = conc, y = dens, type ="h")
plot(x = conc, y = dens, type ="s")
# Add title 
plot(x = conc, y = dens, type ="h",
     col = "red", # colour of line
     lwd = 3, # width of line
     xlab = "Conc", ylab = "Density", # x, y axis labels
     main = "ELISA Assay of DNase") # title for plot
```

**Tip: Open graphics windows before calling plot() to view larger plot/several graphs at once**

Function - Platform
In Windows, windows() or win.graph()
In Unix, X11()
In Mac, quartz() or x11()
```{r}
# windows()
x <- seq(-pi,pi, by = 0.1)
plot(x, sin(x), main="The Sine Function")
```

**Plot line graph with regression  **

- Linear regression: statistical analysis technique used to determine the extent to which there is a linear relationship between a dependent variable and one or more independent variables  
  
- create simple regression using lm() or "linear model" function  
- 2 common parameters:  
      + formula: describes the model with the format “Y-var ~ X-var”, where Y-var is the dependent variable and X-var is independent variable    
      + data: the variable that contains the dataset   
- summary() is used to find the intercept and coefficients (under "Estimate"), R^2 and p-value  
     Recall: y = mx + b, where m is c  
       
Read more: https://www.r-bloggers.com/r-tutorial-series-simple-linear-regression/
```{r}
# Plot 
plot(DNase$conc, y <- DNase$density, 
     type = "p", # line graph
     col = "red", # colour of line
     lwd = 3, # width of line
     xlab = "Concentration", # x axis label
     ylab = "Density", # y axis label
     main = "ELISA assay of DNase") # title of plot
# Make a linear model with the density and concentration variables
fit <- lm (density ~ conc, data = DNase) 
fit
# Get statistics
summary(fit)
# Add a line to the plot (new layer)
abline(fit)
# Add text to the plot (new layer)
# text()
```

**Make a basic histogram of a numeric vector using hist()**

- ref: https://rstudio-pubs-static.s3.amazonaws.com/7953_4e3efd5b9415444ca065b1167862c349.html  
- Histograms: show frequencies for ranges of values 
```{r}
hist(DNase$density)
```

**Saving to File**

1 - Plot by exporting using "Export" in plot panel (bottom-right pane) or in window menu
2 - Plot by opening a file connection
  Format: type_of_file("filename")
    pdf(rplot.pdf): pdf file
    png(rplot.png): png file
    jpeg(rplot.jpg): jpeg file
    postscript(rplot.ps): postscript file
    bmp(rplot.bmp): bmp file
    win.metafile(rplot.wmf): windows metafile
  Note: Use "dev.off()" to close the connection after plotting
```{r}
# Create 2 vectors with 10 elements/values
TEN_Random <- runif(10, 0.0, 1.0)
TEN_sequence <- seq(from = 1, to = 30, length.out = 10)
# Make the second value missing in y 
TEN_sequence[2] <- NA
# Create a "Plots" directory in your current working directory
dir.create("Plots")
# Save to png file
png("Plots/plot.png") # open "device" connection
plot(x = TEN_sequence, y = TEN_Random, xlab = "Index", ylab = "Random values") # plot
dev.off() #close connection
```

**Create an editable graph from R using ReporteRs package**

- Editable vector graphics can be created and saved in a Microsoft document 
- Read here: http://www.sthda.com/english/wiki/create-an-editable-graph-from-r-software

## Practice
The "women" data set in R gives the average heights and weights for American women aged 30 to 39.  
a) Print the first 15 rows to the console. (Hint: use the "n" argument in head() function)  
b) Create a folder called "Data Sets" in your current working directory.  
c) Write the women data frame as a csv file to the Data Sets folder (exclude row names).  
d) Read this file back into R and assign it to a variable called "women.df".  
e) Plot a histogram of the heights column.  
f) Find the mean and standard deviation of heights (Recall: vectors tutorial)  
g) print the variables from f) in a statement "The mean and standard deviation of the heights is __ and __" (Hint: use sprintf() or paste() )  
h) Plot a scatter plot, where x = height and y = weight. Relabel x and y axes to "Height (in)" and "Weight (lbs)" respectively.  
i) Save f) to a jpeg file.  
 
Solution
```{r}
data("women")
# a) Print using head()
head(women, n = 15)
# b) Create a folder using dir.create()
dir.create("Data Sets")
# c) Write to csv using write.csv()
write.csv(x = women, file = "Data sets/women.csv", row.names = F)
# d) Read using read.csv()
women.df <- read.csv(file = "Data sets/women.csv")
# e) Plot histogram using hist()
hist(women.df$height)
# f) Find mean using mean() and standard deviation using sd()
mean.hts <- mean(women.df$height)
sd.hts <- sd(women.df$weight)
# g) Print
sprintf("The mean and standard deviation of the heights is %s and %s", mean.hts, sd.hts)
paste("The mean of the heights is ", mean.hts, " and ", sd.hts, sep = "")
# h) Plot using plot()
plot(x=women.df$height, y=women.df$weight, xlab = "Height (in)", ylab = "Weight (lbs)", main = "Avg Heights and Weights for American Women 30-39")
# i) Save as jpeg
jpeg(filename = "heights_vs_weights.jpeg")
plot(x=women.df$height, y=women.df$weight, xlab = "Height (in)", ylab = "Weight (lbs)", main = "Avg Heights and Weights for American Women 30-39")
dev.off()
```
