---
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
---
# Correlation Analysis
Objective: To learn how to perform correlation analysis in R

Summary:
- cor.test(x,y, method) computes the correlation coefficient and significance between two variables  
- cor() creates correlation matrix from a numeric matrix/data frame
- in corrplot package, cor.mtest() gives p-values and corrplot() plots a graph (i.e. correlogram) of a correlation matrix

**What is correlation test?**

- Correlation test is used to evaluate the association between two or more continuous (numeric) variables.

## Correlation Methods

(Note: Ref for formulas: http://www.sthda.com/english/wiki/correlation-test-between-two-variables-in-r)

Pearson correlation: 
- parametric test, which measures a linear dependence between two variables (x and y)
- Correlation coefficient (r) is a value between -1 and 1:
    -1 - strong negative correlation: x increases, y decreases (\)
    0 - no association between the two variables (x and y) (-)
    1 - strong positive correlation: x increases, y increases (/)

- Kendall tau and Spearman rho are rank-based correlation coefficients (non-parametric)

Compute correlation in R

** R functions**

- cor() computes just correlation coefficient (acts on vectors and data frames)

- cor.test() computes correlation coefficient and p-value
- Arguments:
    x, y = numeric vectors with the same length
    method = correlation method, e.g. "pearson"
- return values:
    p.value = p-value of the test
    estimate = the correlation coefficient


Here, we'll use the built-in R data set mtcars for our analysis
```{r}
# Look at the first 6 rows
head(mtcars, 6)
```

Visualize your data using scatter plots
```{r}
library("ggpubr")
# ggscatter() creates a scatter plot
ggscatter(mtcars, x = "mpg", y = "wt", # specify data and aesthetics 
          add = "reg.line", conf.int = TRUE, # add regression to plot
          cor.coef = TRUE, cor.method = "pearson", #label correlation coeffient
          xlab = "Miles/(US) gallon", ylab = "Weight (1000 lbs)") # label axes
```

Run Pearson correlation test
```{r}
res <- cor.test(mtcars$wt, mtcars$mpg, method = "pearson")
res
# In the result above:
# t is the t-test statistic value (t = -9.559),
# df is the degrees of freedom (df= 30),
# p-value is the significance level of the t-test (p-value = 1.29410^{-10}).
# conf.int is the confidence interval of the correlation coefficient at 95% (conf.int = [-0.9338, -0.7441]);
# sample estimates is the correlation coefficient (Cor.coeff = -0.87).
# 
# Interpretation of result:
# - p-value < 0.05 and r = -0.87 < 0. So wt and mpg are significantly correlated and strongly negatively correlated.
```

Recall: result is a list object
```{r}
# "res" is a list object so we can extract values from the result 
# Extract the p.value
res$p.value
# Extract the correlation coefficient
res$estimate
```

## Correlation Matrix: 
** Analyze, Format and Visualize**

Read more: http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-analyze-format-and-visualize-a-correlation-matrix-using-r-software

Correlation matrix is used to analyze the correlation between multiple variables at the same time (performs pairwise correlation tests)

**Draw scatter plots using chart.Correlation()**

The function chart.Correlation() can be used to display multiple charts for a correlation matrix.
```{r}
# # # Install package
# # install.packages("PerformanceAnalytics")
# # Load package into environment
# library("PerformanceAnalytics")
# # Subset data frames (select certain columns)
# my_data <- mtcars[, c(1,3,4,5,6,7)]
# # Plot scatter plots, histograms and correlation using chart.Correlation()
# chart.Correlation(my_data, histogram=TRUE, pch=19)
```

**Draw correlogram using corrplot()**

- Correlogram is a graph of correlation matrix
```{r}
# Create correlation matrix
M <- cor(mtcars, method = "pearson")
# Print the first 6 rows, round all values to 2 decimal points
head(round(M,2))
```

- use corrplot() to plot the graph of the correlation matrix

- Arguments
    corr = correlation matrix 
    method = visualization: circle, square, ellipse, number, shade, color, pie
    ** many more options to customize: https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html
```{r}
# # Install and load package
# install.packages("corrplot")
# Plot correlogram
library(corrplot)
corrplot(M, method="circle")
corrplot(M, method="color")

# Positive correlations = blue and negative correlations = red
# Color intensity and size of the circle are proportional to the correlation coefficients

# Display the correlation coefficient:
corrplot(M, method="number")
```

- Use cor.mtest() to make matrix of p-value
```{r}
# Create a list of matrices of p-value, lower confidence interval (CI) and upper CI for a correlation matrix
p.mat <- cor.mtest(mtcars)
# Get the p-values from the list object
p.mat <- p.mat$p
# Print the first 6 rows and first 5 columns
head(p.mat[, 1:5])

# Plot correlogram with p-values
# Specialized the insignificant value according to the significant level
corrplot(M, type="upper", p.mat = p.mat, sig.level = 0.01, order="hclust")

# Note: correlations with p-value > 0.01 are considered as insignificant (i.e. correlation with crosses)
# type = upper means show only the upper triangle of the matrix
# order = hclust means order the variables by hierarchal clustering
```


## Practice: Correlation analysis
Motor trend car road tests (mtcars)
Fuel consumption and 10 additional aspects (variables) of automobile design tested in 1974 are given in an in-built R dataset called "mtcars".  
a) Look at the structure of mtcars. (Use str())  
b) Is the first column/variable numeric?  
c) Are the variables "mpg" and "drat" in mtcars normally distributed? (can use shapiro.test())  
d) What is the Pearson correlation coefficient and p-value of the correlation between mpg and drat?  
e) Is the correlation from d) positive or negative? Is it significant?  
f) Plot a scatter plot of the two variables. Bonus: Add a line for regression. (use plot, ggplot, ggscatter from ggpubr)  
g) Create a Pearson correlation matrix of mtcars. Assign it to a variable "cor_mat"  
h) Create a p-value matrix for the correlation matrix in g). Assign it to a variable "p_mat".  
i) Create a correlogram of the correlation matrix and p-values in g and h using the corrplot package.  