---
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
---
# Dataframe Manipulation
Objective: To reformat data frames using tidyr and learn how to pipe using magrittr


Quick Review of vectors and data frames
```{r}
# Store data/parameters in objects in R by assigning it to a variable using: variable_Name <- value

# Make a "character" vector called person using function c()
people <- c("Tinky", "Dipsy", "Lala", "Po")
# Assign names to vector using a character vector of same length
names(people) <- c("purple", "green", "yellow", "red")
# Note () brackets reserved for functions

# Access data in vectors using [] 
people[c(1,3)] #use indexing 1st and 3rd elements
people[1:3] #1st, 2nd and 3rd person
people[-2] #negative indexing to get all values except the 2nd
people[c(T,T,F,F)] #use logical vector to get 1st and 2nd elements
people[c("purple", "yellow")] #use character vector to get 1st and 3rd elements *for named vectors

# Make a dataframe or table called "df" to store vectors of different data types
# Assign column names for each variable using = 
df <- data.frame(Person=people, 
                 Height=seq(from=180, to=120, length.out = 4))

# Access data in df using [rows_of_interest, cols_of_interest]
df[1,] #1st row
df[,1] #1st column
df[,c(T,F)] #1st column
df[,"Person"] #1st column
df$Person #1st column

# Modify variables in dataframe by re-assignment
df$Height <- df$Height + 2 #add 2 to all heights
```

**Today's tutorial:**

Original tutorials: 
- tidyr: http://www.sthda.com/english/wiki/tidyr-crucial-step-reshaping-data-with-r-for-easier-analyses
- magrittr pipe: https://www.datacamp.com/community/tutorials/pipe-r-tutorial
- dplyr pipe: https://uoftcoders.github.io/studyGroup/lessons/r/intro/lesson/


## Reshape Data 
- using tidyr package
- ways to manipulate dataframes to facilitate organization and improve readability of code

Useful functions:  
  gather(): gather (collapse) columns into rows
  spread(): spread rows into columns
  separate(): separate one column into multiple using "delimiter"
  unite(): unite multiple columns into one using "delimiter"

```{r}
# # Install package
# install.packages("tidyr")
# Load package into environment to use
library("tidyr")
```

### Data
- Take a subset of built-in "USArrests" data sets (state x violent crimes)
```{r}
head(USArrests)
# Create a new data frame with just 4 of the observations and remove 4th column
df <- USArrests[c(5,12,32,43), -4]
# Add a column to beginning of data frame with state names = row names
df <- cbind(state = rownames(df), df)
# Remove row names
rownames(df) <- NULL
```             

**gather()**: collapse columns into rows

- collapses multiple columns into key-value pairs
- converts data format "wide" to long
- alternative of melt() function [in reshape2 package]
- useful in ggplot, applying a function (ex. p_adjust() to multiple samples at once)

gather(data, key, value, ...)

Arguments: 
data = data frame 
key, value = Names of key and value columns to create in output
value = Specification of columns to gather. Allowed values are:
    variable names
    if you want to select all variables between a and e, use a:e
    if you want to exclude a column name y use -y
    for more options, see: dplyr::select()

Examples of usage:
Gather all columns except the column state (-)
```{r}
gather(df, key = "arrest_variable", #this new column will be your column names
       value = "arrest_rate", #values
       -state)
```

Gather only Murder and Assault columns
```{r}
gather(df, key = "arrest_variable", 
       value = "arrest_rate", 
       Murder, Assault)

```
Note that, the two columns Murder and Assault have been collapsed and the remaining columns (state, UrbanPop and Rape) have been duplicated.

Gather all variables between Murder and UrbanPop
```{r}
df2 <- gather(df,
              key = "arrest_variable",
              value = "arrest_rate",
              Murder:UrbanPop)
```

**spread()**: spread two columns into multiple columns

- spread() does the opposite of gather(). It takes two columns (key and value) and spreads into multiple columns
- produces a wide data format from a long one 
- alternative of the function cast() [in reshape2 package]

spread(data, key, value)

Arguments: 
- key = name of the column whose values will be used as column headings.
- value = name of the column whose values will populate the cells.

```{r}
# Spread df to turn back to the original data
spread(df2, 
       key = "arrest_variable", 
       value = "arrest_rate")
```      

**unite()**: Unite multiple columns into one

- takes multiple columns and paste them together into one.

unite(data, col, ..., sep = "_")

Arguments: 
    col = The new (unquoted) name of column to add.
    sep = Separator to use between values
```{r}
# unites the columns Murder and Assault
df4 <- unite(df, 
             col = "Murder_Assault",
             Murder, Assault,
             sep = "_")
df4
```

**separate()**: separate one column into multiple

- seperate() does the opposite of unite()
- takes values inside a single character column and separates them into multiple columns

separate(data, col, into, sep = "[^[:alnum:]]+")

Arguments: 
    col = Unquoted column names
    into = Character vector with names of new columns to be created
    sep = Separator between columns:
          If character, is interpreted as a regular expression  
          If numeric, interpreted as positions to split at  
```{r}
# Separate the column Murder_Assault [in df4] into two columns Murder and Assault:
separate(df4,
         col = "Murder_Assault",
         into = c("Murder", "Assault"),
         sep = "_")
```

Alternative forms: gather_(), spread_(), unite_()
- useful when making your own functions
```{r}
# Make variable names (ie. would be arguments in your functions)
col_names <-  c("Murder", "Assault")
key_column <- "arrest_variable"
value_column <- "arrest_rate"

# opposite functions:
# alt to gather() - wide to long format
df2 <- gather_(df,
               key_col = key_column,
               value_col = value_column,
               gather_cols = col_names) # define columns to gather
# alt to spread() - long to wide format
spread_(df2, 
       key = "arrest_variable",
       value = "arrest_rate")

# opposite functions:
# alt to unite() - unite many columns into one
df4 <- unite_(df,
              col = "Murder_Assault",
              from = c("Murder", "Assault"),
              sep = "_")
# alt to seperate() - seperate one column into multiple
separate_(df4,
           col = "Murder_Assault",
           into = c("Murder", "Assault"),
           sep = "_")
```

## Pipe operator 
- in R it is... %>%
- pipes are used to chain multiple operations
- similar to how the Bash shell | pipe works
- forward-pipe operator is %>% (in magrittr and dplyr packages)
```{r}
# install.packages("magittr")
library(magrittr)
```

- x %>% f is equivalent to f(x).
- Read more: https://www.datacamp.com/community/tutorials/pipe-r-tutorial

- advantages: decrease development time and to improve readability and maintainability of code

Example 1
```{r}
# First, df is passed to gather() function
# Next, the output of gather() is passed to some functions for transformations.. 
# Then, we convert to its original data format by unite() function
new_df <- 
  df %>% 
  # convert to long format
  gather(key = "arrest_variable",
         value = "arrest_rate",
         Murder:UrbanPop) %>%
  # Apply some transaformations here
  #..
  # return to original wide format
  unite(col = "attribute_estimate",
        "arrest_variable", "arrest_rate")

# note: we don't keep mentioning the "df" object because that is the data being passed on 
```

Example 2:
```{r}
# Manipulate data, but combine operations using pipe
new_df <- 
  df %>%
  subset(Assault > 200) %>% 
  transform(AttemptedMurder = Murder %>% multiply_by(0.4251)) %>% 
  print %>% 
  plot


# # Explanation ----
# new_df <- # final output will be assigned to this variable
#   df %>% # pass in original data frame.. 
#   subset(Assault > 200) %>% # subset so Assault column > 200
#   transform(AttemptedMurder = Murder %>% multiply_by(0.4251)) %>% # add a new column 
#   print %>% # print to console
#   boxplot # visualize and save plot 
```


## Practice
The built-in R dataset "ChickWeight" provides experimental data on the effect of diet on early growth of chicks.

a) Look at the structure of the data (str()). How many rows are there?
b) This data is presented in a "long" format. Spread the counts data across columns in "wide" format specified by the key "Time" and value "weight". Save output a variable "df".
c) Since the new columns are the timepoints of the experiment ("0:21"), unite these back into a single column called "Timepoint". Save in a variable called df2.
d) Do b) and c) in a single line of code using a pipe operator. That is, make df2 without producing df.


Solution
```{r}
# # a) use str()
# str(ChickWeight)
# # To find number of rows, look at how many observations there are (obs.)
# # or use nrow() or dim()
# nrow(ChickWeight)
# dim(ChickWeight)
# # ANSWER: 578
# 
# # b) use spread()
# library(tidyr)
# df <- spread(ChickWeight, key = "Time", value = "weight")
# 
# # c) unite
# # time_cols <- colnames(df)[3:ncol(df)]
# df2 <- unite(df, col="Timepoint", c("0", "2", "4", "6", "8", "10", "12", "14", "16", "18", "20", "21"), sep = "_") 
# 
# # d) use %>% from dplyr
# library(magrittr)
# df2 <- ChickWeight %>%
#         spread(key = "Time", value = "weight") %>%
#         unite(col="Timepoint", c("0", "2", "4", "6", "8", "10", "12", "14", "16", "18", "20", "21"), sep = "_")

```