---
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
---
# Repetitive Jobs
Objective: To learn about tools to prevent copying and pasting the same code to replicate instructions

We will cover:  
- how to use functions (review)
- make your own function(){}
- loops 
- vectorization - apply family (apply(), lapply())

## Functions
Recall: A quick note on functions

- Functions are a group of statements that together perform a specific task
- Functions have a name, e.g. setwd, getwd
- We can "call" functions (also called commands in this context) in R for convenience (so we don't have to rewrite the code)
  
- Use ? or help() command To find information for a particular function  
```{r}
?print
help(print)
```

- Functions have the format: function_name( arguments )
- "Arguments" are required or optional parameters used by the function to accomplish the action
- Functions can "return" a value (or not)
```{r}
# Type a comma and press tab to see arguments (function parameters) in RStudio
print(x = "Hello world")
print(x = 33.9431249, digits = 4)
```

**How to define a function:**

   name_of_function <- function(list_of_arguments){
                          # code to do what function wants
                      }

- arguments/args: names of variables you will use in your function (seperated by commas, within parantheses)
- return value: the value that is returned by a function (optional) - use return()

**Make a function using function()**

Make a function that adds 2 values, x and y
```{r}
add <- function(x, y) { #arguments within parantheses
  x.y.sum <- x + y    # body of function within {}
  return(x.y.sum)     # return value specified in return()
}
```

Call your function the same way you use any other.
```{r}
add(4,5) #returns 9; if not saved to a variable, it will print
# add(4) # Error in add(4) : argument "y" is missing, with no default
```

Automatic Returns - In R, it is not necessary to include the return statement. R automatically returns whichever variable is on the last line of the body of the function.

**Default arguments**
Set defaults if not specified by user. 
```{r}
# Define a function
add <- function(x, y=0) { # x is required argument, but if y is not specified, it is 0
  x.y.sum <- x + y    
  return(x.y.sum)    
}
# Use function
add(4,5)
add(4)
```

Define a function lb.to.kg that converts weights from pounds (lb) to kilogram (kg):
```{r}
# for an approximate result, divide the mass value by 2.205
lb.to.kg <- function(wt_lb) { #argument within parantheses
  wt_kg <- wt_lb/2.205      # body of function within {}
  wt_kg <- round(wt_kg, digits = 3)
  return(wt_kg)           # return value specified in return()
}

# Let's try running our function. Calling our own function is no different from calling any other function:
lb.to.kg(32)
lb.to.kg(212)
```

Summary:
- Define a function using name <- function(args...) { body... }
- Call a function using name(values.. )

- taste of how larger programs are built: we define basic operations, then combine them in ever-larger chunks to get the effect we want. 

## Loops

Based on following tutorial:  https://swcarpentry.github.io/r-novice-inflammation/03-loops-R/index.html 

Suppose we want to print each word in a sentence. One way is to use six print statements:
```{r}
# Make a character vector
sentence1 <-  c("Try", "printing", "this")

# Make a function to print each value in sentence vector
print_sentence <- function(sentence) {
  print(sentence[1])
  print(sentence[2])
  print(sentence[3])
}
# Call function
print_sentence(sentence1)
```

However this function won't work as intended (doesn't scale and fragile) if there are more than 3 values in sentence argument.
```{r}
# Make another sentence vector
sentence2 <- c("Let", "the", "computer", "do", "the", "work")
# Call function
print_sentence(sentence2)
```

Or if the sentence is shorter, NAs (missing values) are introduced.
```{r}
# Make another sentence vector
sentence3 <- c("Try", "this")
# Call function
print_sentence(sentence3)
```

Loop 
- All modern programming languages provide special constructs that allow for the repetition of instructions or blocks of instructions.

- There are 3 main types of loops in R: for, while, repeat
- You can add break; and next; to skip a iteration if it does not passed a test (ie. condition) 

Read more: https://www.datacamp.com/community/tutorials/tutorial-on-loops-in-r

**The 'for' loop construct**


    for (variable in collection) {
      # do things with variable
    }

Eg. make a for loop
```{r}
# Make a for loop to print out each element in sentence1
for (word in sentence1){ # word will take on the value of "sentence1" element by element until there are no elements left
  # print the current word
  print(word)
}
```

Add a condition
```{r}
# Make a for loop to print out each element in sentence1
for (word in sentence1){ # word will take on the value of "sentence1" element by element until there are no elements left
  # Skip current iteration if the current word is equal to  "printing"
  if(word == "printing"){
    next;
  }
  # print the current word
  print(word)
}
```
Eg. make a while loop
```{r}
# Intialize an index counter 
i <- 1 

# Make a while loop to print out each element in sentence1
# Note: length() is the number of elements in a vector or list
while (i <= length(sentence1)){ # i will test this condition and run the code within {} until it is FALSE
  # Define word variable by indexing sentence1
  word <- sentence1[i]
  # print the current word
  print(word)
  # Update counter by 1 - i.e. move to the next index - must do this or you'll end up with an infinite loop!
  i <- i+1
}
```

Applications: 
- processing/analyzing multiple files (look at linked tutorial)
- plotting graphs from columns in same dataframe
```{r}
# eg. Use a for loop to apply same preprocessing steps to multiple csv files
for (filename in list.files(pattern="csv")){ 
  # Read csv file into a variable called df
  df <- read.csv(filename)
  
  if(is.null(df)){
    next;
  }
  
  # downstream processing df
}
```

## Vectorization
- Alternative to looping 
- Vectorization is a feature of R that allows you to apply an operation to data/vectors at the same time
- This makes code more readable
- most functions in R perform vectorization intrinsically, such as colSums(), rowMeans(), and even basic arithmetic operations (e.g +, -, *)

e.g. Power of vectorization: Add 2 to all values
```{r}
# Define a numeric vector called values
values <- c(3,5,6,10)

# In many programming languages you may have to run a loop to do this
for(value in values){
  new_value <- value + 2
  print(new_value)
}

# But R lets you add 2 to all values at once - this is called vectorization
values + 2
```

## Intro to apply 
- apply family functions utilize the concept of vectorization
- functions: apply(), sapply(), lapply(), mapply(), rapply(), tapply(), vapply()
- within these commands, you may apply a function or operation to the values 
- put name of function into FUN (no parantheses) or make your own


- most frequently used are:

1) lapply ("list" apply)
- applies a function on 1D data - list or vector
- returns a list
- note: use unlist() to convert the resulting list to vector
```{r}
# e.g. print vector elements
lapply(sentence1, FUN = print)
# e.g. add 2 to each element by making your own function
lapply(values, FUN = function(value){
  value + 2
})
```

2) apply
- apply a function to 2D data - matrix or data frame
- specify if function should be apply on rows or columns (using MARGIN argument: 1 = row, 2 = column)
```{r}
# Read input table
df <- read.delim(file = "07-Casey-mamm-mouse-proteome-sample.txt", row.names = 1)

# Get standard deviation of columns
apply(df, MARGIN = 2, FUN = sd)

# Transform data to z-scores (which account for sd of each row)
z_scores <- apply(df, MARGIN = 1, FUN = function(x){ # in this function, x is a numeric vector of each row
  (x-mean(x))/sd(x)
})
# Look at first 6 rows and 6 columns
z_scores[1:6, 1:6]
# Note apply() has transposed our original matrix; re-transpose using t()
z_scores <- t(z_scores)
```


## Practice
Converting cm to in
a) Define a function called "cm_to_in". This function takes in a numeric variable and converts it from centimeters to inches by dividing the length value by 2.54. Round to 2 significant digits using round(). Return the result. 
b) Test your function on any number.
c) Make a numeric vector of 5 different values called "cm_measurements".
d) In a loop, 
- convert the cm vector to inches using your function in a) and save to variable called "in_inches"
- print "in_inches" to console using print()
e) Use the lapply() function to apply your function to each element.
f) Pass your vector into your function. What's the result? This is called vectorization.

Solution
```{r}
# a) make a function
cm_to_in <- function(value){
  # Divide value by 2.54
  result <- value/2.54
  # Round to 2 digits
  result <- round(result, digits = 2)
  return(result)
}
# b) pass in single numeric value to function
cm_to_in(45)
# c) make a numeric vector
cm_measurements <- c(35,63,53,67,32)
# d)  make for loop
for(in_cm in cm_measurements){
  # Save output of function to variable
  in_inches <- cm_to_in(in_cm)
  # Print
  print(in_inches)
}
# e) use lapply() on vector
lapply(cm_measurements, FUN = cm_to_in)
# f) pass in whole vector to function
cm_to_in(cm_measurements)
```


